Real-time resolution of short-read assembly graph using ONT long reads

### 2020-09-13 plos_latex_clean.tex
reorder algorithm1&2

We also provide a Graphical User Interface (GUI) for \npgraph{}. The GUI includes the dashboard for controlling the settings of the program and a window for visualization of the assembly graph 
in real-time (Figure~\ref{figure:npgraph_gui}). 
In this interface, the assembly graph loading stage is separated from the actual assembly process so that users can check for the graph quality first before carry out any further tasks.
A proper combination of command line and GUI can provide an useful streaming pipeline that copes well with MinION output data. The practice is to support the real-time monitoring of
the results from real-time sequencing~\cite{CaoGC2016,Cao2017scaffolding,Nguyen2017barcode} that allow the analysis to take place abreast to a nanopore sequencing run.

@@

%\subsection*{Untangling assembly graph by stream of nanopore data.}
\subsubsection*{Building bridges in real-time}
Bridge is the data structure designed for tracking the possible connections between two anchored nodes (of unique contigs) in the assembly graph.
A bridge must start from a unique contig, or \emph{anchor} node, and end at another when completed. Located in-between are nodes known as \emph{steps} and distances between them are called \emph{spans} of the bridge. Stepping nodes are normally repetitive contigs and indicative for a path finding operation later on. In a complicated assembly graph, the more details the bridge, \emph{a.k.a.} more steps in-between, the faster and more accurate the linking path it would resolve. A bridge's function is complete when it successfully return the ultimate linking path between 2 anchors.

@@

\subsubsection*{Path finding algorithm}
Given a bridge with 2 anchors, a path finding algorithm is invoked to find all candidate paths between them. Each of these paths is given a score of alignment-based likelihood which are updated immediately as long as there is an appropriate long read being generated by the sequencer. As more nanopore data arrives, the divergence between candidates' score becomes greater and only the top-scored ones are kept for the next round.
We implement a modified stack-based version utilizing Dijkstra's shortest path finding algorithm~\cite{Dijkstra1959} to reduce the search space when using Depth-First Search.

Due to false positive alignments from shorter contigs to the long reads, not all of the reported step nodes would appear in the ultimate path resolved by the bridge. 
In most cases, the accumulated score of each step indicates its likelihood to be the true component of the final solution.
For that reason, a strategy similar to binary search is employed to find a path across 2 anchors of a bridge as shown in Algorithm~\ref{algo:binsearch}. Briefly, it recursively looks for sub-paths connecting pairs of step nodes in-between 2 anchors and return a final continuous path with the highest score.
We define Algorithm~\ref{algo:findpath} as the \emph{base} sub-rountine to find paths between a \emph{connectable} pair of two step nodes given their estimated distance. In which, function 
\[
\mathtt{shortestTree}(\overrightarrow{vertex},distance) : (V,Z) \rightarrow V^n
\]
from line 3 builds a shortest tree rooted from $\overrightarrow{v}$, following its direction until a distance of approximately $d$ (with a tolerance for the nanopore error rate) is reached. This task is implemented based on Dijkstra algorithm.
This tree is used on line 4 and in function $includedIn()$ on line 19 to filter out any node or edge with ending nodes that do not belong to the tree.

@@

\begin{algorithm}[!hpt]
\DontPrintSemicolon
\KwData{Assembly graph $G\{V,E\}$}
\KwIn{Brigde $B:\lbrace\overrightarrow{v_0}, ...\overrightarrow{v_k}, ...,\overrightarrow{v_n}\rbrace$ with $\overrightarrow{v_0}$ and $\overrightarrow{v_n}$ are two anchors, $\{\overrightarrow{v_k}\}, k=1\ldots(n-1)$ are steps in-between}
\KwOut{Set of candidate paths connecting $\overrightarrow{v_0}$ to $\overrightarrow{v_n}$ that maximize the likelihood of the step list.}
\SetKwFunction{BinaryBridging}{BinaryBridging} 
\SetKwProg{Fn}{Function}{:}{}
\Fn{\BinaryBridging{B}}{
\tcc{search for the contig with maximum score from the step list (two ends excluded)}
$m$:=$\argmax_k(\overrightarrow{v_k}.score())$ \;
\tcc{if there is no step in-between, run path finding algorithm above directly and return the result}
\If{$M.size() \equiv 2$}{ 
	\Return{$\mathtt{DFS}(B.start(),B.end().B.distance())$}
}
\tcc{divide the original bridge $B$ into 2 bridges by $v_m$: $BL$ and $BR$}
$BL$:=$\lbrace B.start(), ...,\overrightarrow{v_m}\rbrace$ \;
$BR$:=$\lbrace \overrightarrow{v_m}, ...,B.end()\rbrace$ \;
\tcc{Return the join of running recursive function on two sub-bridges}
\Return{$\BinaryBridging(BL)\Join\BinaryBridging(BR)$} 
}
\caption{Recursive binary bridging to connect 2 anchor nodes.}
\label{algo:binsearch}
\end{algorithm}

>>>>

We also provide a Graphical User Interface (GUI) for \npgraph{}. The GUI includes the dashboard for controlling the settings of the program and a window for visualization of the assembly graph 
in real-time (Figure~\ref{figure:npgraph_gui}). 
In this interface, the assembly graph loading stage is separated from the actual assembly process so that users can check for the graph quality first before carry out any further tasks.
A proper combination of command line and GUI can provide an useful streaming pipeline that copes well with MinION output data. This is designed to support the real-time monitoring of
the results from real-time sequencing~\cite{CaoGC2016,Cao2017scaffolding,Nguyen2017barcode} that allow the analysis to take place abreast to a nanopore sequencing run.

@@

%\subsection*{Untangling assembly graph by stream of nanopore data.}
\subsubsection*{Building bridges in real-time}
Bridge is the data structure designed to identify the possible connections between two anchored nodes (of unique contigs) in the assembly graph.
A bridge must start from a unique contig, or \emph{anchor} node, and end at another if completed. Located in-between are nodes known as \emph{steps} and distances between them are called \emph{spans} of the bridge. Stepping nodes are normally repetitive contigs and indicative for a path finding operation later on. In a complicated assembly graph, the more details the bridge, \emph{a.k.a.} more steps in-between, the faster and more accurate the linking path it would resolve. A bridge's function is complete when it successfully return the ultimate linking path between 2 anchors.

@@

\subsubsection*{Path finding algorithm}
Given a proposed bridge with 2 anchors $B=\lbrace\overrightarrow{v_0}\,..., \overrightarrow{v_n}\rbrace$ identified from previous step, a searching algorithm is implemented to find the final bridging solution, \IE{} an ungapped path among all possiblities. To do so, each of the candidates is given a score of alignment-based likelihood which are updated immediately as long as there is an appropriate long read being generated by the sequencer. As more nanopore data arrives, the score divergence increases and only the highest one would be selected for the next round.

Overall, the algorithm employs a binary search strategy as shown in Algorithm~\ref{algo:binsearch}. Starting with the original bridge $\lbrace\overrightarrow{v_0}\,..., \overrightarrow{v_n}\rbrace$, we recursively break the target bridge in half at one of its \emph{step} $\overrightarrow{v_k}$ in-between then find the sub-paths of $\lbrace\overrightarrow{v_0}\,..., \overrightarrow{v_k}\rbrace$ and $\lbrace\overrightarrow{v_k}\,..., \overrightarrow{v_n}\rbrace$. Algorithm~\ref{algo:binsearch} will call Algorithm~\ref{algo:findpath} when the base condition is met, meaning the target bridge is unbreakable (no more in-between \emph{step} available).

\begin{algorithm}[!hpt]
\DontPrintSemicolon
\KwData{Assembly graph $G\{V,E\}$}
\KwIn{Brigde $B:\lbrace\overrightarrow{v_0}, ...\overrightarrow{v_k}, ...,\overrightarrow{v_n}\rbrace$ with $\overrightarrow{v_0}$ and $\overrightarrow{v_n}$ are two anchors, $\{\overrightarrow{v_k}\}, k=1\ldots(n-1)$ are steps in-between}
\KwOut{Set of candidate paths connecting $\overrightarrow{v_0}$ to $\overrightarrow{v_n}$ that maximize the likelihood of the step list.}
\SetKwFunction{BinaryBridging}{BinaryBridging} 
\SetKwProg{Fn}{Function}{:}{}
\Fn{\BinaryBridging{B}}{
\tcc{search for the contig with maximum score from the step list (two ends excluded)}
$m$:=$\argmax_k(\overrightarrow{v_k}.score())$ \;
\tcc{if there is no step in-between, run Algorithm~\ref{algo:findpath} and return the result}
\If{$M.size() \equiv 2$}{ 
	\Return{$\mathtt{DFS}(B.start(),B.end().B.distance())$}
}
\tcc{divide the original bridge $B$ into 2 bridges by $v_m$: $BL$ and $BR$}
$BL$:=$\lbrace B.start(), ...,\overrightarrow{v_m}\rbrace$ \;
$BR$:=$\lbrace \overrightarrow{v_m}, ...,B.end()\rbrace$ \;
\tcc{Return the join of running recursive function on two sub-bridges}
\Return{$\BinaryBridging(BL)\Join\BinaryBridging(BR)$} 
}
\caption{Recursive binary bridging to connect 2 anchor nodes.}
\label{algo:binsearch}
\end{algorithm}

In Algorithm~\ref{algo:findpath}, we implement a modified stack-based Depth-First Search (DFS) using Dijkstra's shortest path finding algorithm~\cite{Dijkstra1959} to reduce the search space.
In which, function 
\[
\mathtt{shortestTree}(\overrightarrow{vertex},distance) : (V,Z) \rightarrow V^n
\]
from line 3 builds a shortest tree rooted from $\overrightarrow{v}$, following its direction until a distance of approximately $d$ (with a tolerance for the nanopore error rate) is reached, similar to the Dijkstra algorithm.
This tree is used on line 4 and in function $includedIn()$ on line 19 to filter out any node or edge with ending nodes that do not belong to the tree.

@@

### 2020-09-13 plos_latex_edit.tex
reorder algorithm1&2

\subsubsection*{Path finding algorithm}
Given a bridge with 2 anchors, a path finding algorithm is invoked to find all candidate paths between them. Each of these paths is given a score of alignment-based likelihood which are updated immediately as long as there is an appropriate long read being generated by the sequencer. As more nanopore data arrives, the divergence between candidates' score becomes greater and only the top-scored ones are kept for the next round.
We implement a modified stack-based version utilizing Dijkstra's shortest path finding algorithm~\cite{Dijkstra1959} to reduce the search space when using Depth-First Search.

Due to false \addnew{positive} alignments from shorter contigs to the long reads, not all of the reported step nodes \cancel{are necessary to be appeared} \addnew{would appear} in the ultimate path resolved by the bridge. 
In most cases, the accumulated score of each step indicates its likelihood to be the true component of the final solution.
For that reason, a strategy similar to binary search\cancel{ing} is employed to find a path across 2 anchors of a bridge as shown in Algorithm~\ref{algo:binsearch}. \addnew{Briefly, it recursively looks for sub-paths connecting pairs of step nodes in-between 2 anchors and return a final continuous path with the highest score.}
\cancel{Before that, }We define Algorithm~\ref{algo:findpath} \addnew{as the \emph{base} sub-rountine} to \cancel{demonstrates the path finding algorithm for} \addnew{find paths between a \emph{connectable} pair of} two \addnew{step} nodes given their estimated distance. In which, function 
\[
\mathtt{shortestTree}(\overrightarrow{vertex},distance) : (V,Z) \rightarrow V^n
\]
from line 3 \cancel{of the algorithm's pseudo code} builds a shortest tree rooted from $\overrightarrow{v}$, following its direction until a distance of approximately $d$ (with a tolerance \cancel{regarding nanopore read} \addnew{for the nanopore} error rate) is reached. This task is implemented based on Dijkstra algorithm.
This tree is used on line 4 and in function $includedIn()$ on line 19 to filter out any node or edge with ending nodes that do not belong to the tree.

>>>>

\subsubsection*{Path finding algorithm}
\cancel{Given a bridge with 2 anchors, a path finding algorithm is used to find all candidate paths between them. Each of these paths is given a score of alignment-based likelihood which are updated immediately as long as there is an appropriate long read being generated by the sequencer. As more nanopore data arrives, the divergence between candidates' score becomes greater and only the top-scored ones are kept for the next round.}
\cancel{We implement a modified stack-based version utilizing Dijkstra's shortest path finding algorithm to reduce the search space when using Depth-First Search.}

\cancel{Due to false alignments from shorter contigs to the long reads, not all of the reported step nodes are necessary to be appeared in the ultimate path resolved by the bridge. 
In most cases, the accumulated score of each step indicates its likelihood to be the true component of the final solution.
For that reason, a strategy similar to binary searching is employed to find a path across 2 anchors of a bridge as shown in Algorithm~2.
Before that, we define Algorithm~1 to demonstrates the path finding algorithm for two nodes given their estimated distance.}

\addnew{Given a proposed bridge with 2 anchors $B=\lbrace\overrightarrow{v_0}\,..., \overrightarrow{v_n}\rbrace$ identified from previous step, a searching algorithm is implemented to find the final bridging solution, \IE{} an ungapped path among all possiblities. To do so, each of the candidates is given a score of alignment-based likelihood which are updated immediately as long as there is an appropriate long read being generated by the sequencer. As more nanopore data arrives, the score divergence increases and only the highest one would be selected for the next round.}

\addnew{Overall, the algorithm employs a binary search strategy as shown in Algorithm~\ref{algo:binsearch}. Starting with the original bridge $\lbrace\overrightarrow{v_0}\,..., \overrightarrow{v_n}\rbrace$, we recursively break the target bridge in half at one of its \emph{step} $\overrightarrow{v_k}$ in-between then find the sub-paths of $\lbrace\overrightarrow{v_0}\,..., \overrightarrow{v_k}\rbrace$ and $\lbrace\overrightarrow{v_k}\,..., \overrightarrow{v_n}\rbrace$. Algorithm~\ref{algo:binsearch} will call Algorithm~\ref{algo:findpath} when the base condition is met, meaning the target bridge is unbreakable (no more in-between \emph{step} available).}

\begin{algorithm}[!hpt]
\DontPrintSemicolon
\KwData{Assembly graph $G\{V,E\}$}
\KwIn{Brigde $B:\lbrace\overrightarrow{v_0}, ...\overrightarrow{v_k}, ...,\overrightarrow{v_n}\rbrace$ with $\overrightarrow{v_0}$ and $\overrightarrow{v_n}$ are two anchors, $\{\overrightarrow{v_k}\}, k=1\ldots(n-1)$ are steps in-between}
\KwOut{Set of candidate paths connecting $\overrightarrow{v_0}$ to $\overrightarrow{v_n}$ that maximize the likelihood of the step list.}
\SetKwFunction{BinaryBridging}{BinaryBridging} 
\SetKwProg{Fn}{Function}{:}{}
\Fn{\BinaryBridging{B}}{
\tcc{search for the contig with maximum score from the step list (two ends excluded)}
$m$:=$\argmax_k(\overrightarrow{v_k}.score())$ \;
\tcc{if there is no step in-between, run Algorithm~\ref{algo:findpath} and return the result}
\If{$M.size() \equiv 2$}{ 
	\Return{$\mathtt{DFS}(B.start(),B.end().B.distance())$}
}
\tcc{divide the original bridge $B$ into 2 bridges by $v_m$: $BL$ and $BR$}
$BL$:=$\lbrace B.start(), ...,\overrightarrow{v_m}\rbrace$ \;
$BR$:=$\lbrace \overrightarrow{v_m}, ...,B.end()\rbrace$ \;
\tcc{Return the join of running recursive function on two sub-bridges}
\Return{$\BinaryBridging(BL)\Join\BinaryBridging(BR)$} 
}
\caption{Recursive binary bridging to connect 2 anchor nodes.}
\label{algo:binsearch}
\end{algorithm}

\addnew{In Algorithm~\ref{algo:findpath}, we implement a modified stack-based Depth-First Search (DFS) using Dijkstra's shortest path finding algorithm~\cite{Dijkstra1959} to reduce the search space.}
In which, function 
\[
\mathtt{shortestTree}(\overrightarrow{vertex},distance) : (V,Z) \rightarrow V^n
\]
from line 3 builds a shortest tree rooted from $\overrightarrow{v}$, following its direction until a distance of approximately $d$ (with a tolerance for the nanopore error rate) is reached, similar to the Dijkstra algorithm.
This tree is used on line 4 and in function $includedIn()$ on line 19 to filter out any node or edge with ending nodes that do not belong to the tree.
